Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-17 model.tex
rewrote the ARIMA section without a zillion equations

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $px_i$ at time $i$ is predicted to
be $$px_i = x_{i-1}$$ The \naive ~prediction strategy uses an average
to generate the forecast: the prediction $px_i$ is simply the running
average of observervations up to that point: $$px_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometric techniques and prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are volatile, noisy, and possess very little predictive
structure, but they do not---on the average---vary a great deal, so
the random-walk method's strategy of simply guessing the last known
value is not a bad choice.  If a signal has a unimodal distribution
with low variance, the \naive ~prediction strategy will perform quite
well---even if the signal is highly complex---simply because the mean
is a good approximation of the future behavior.  Also, a temporal
average effects a low-pass filtering operation.  In a signal with very
little predictive structure, then, the \naive ~prediction strategy
effectively mitigates the complexity.

Both of these methods have significant weaknesses, however, because
they do not model the temporal patterns of the time series or even the
distribution of its values.  Thus, they cannot track changes in that
structure, which causes them to fail in a number of important
situations.  Random-walk strategies are a particularly bad choice for
time series that change drastically at every time step.  In the worst
case---a large-amplitude square wave whose period is equivalent to
twice the sample time---a random-walk prediction would be exactly 180
degrees out of phase with the true continuation.  The \naive ~method
would be a better choice in this situation, since it would always
predict the mean.  It would, however, fail when a signal has a number
of long-lived regimes that have significantly different means.  In
this situation, the inertia of the \naive ~method's accumulating mean
is a liability and the agility of the random-walk method is an
advantage, since it can respond quickly to regime shifts.

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The ARIMA and LMA
methods introduced in Sections~\ref{sec:lma} and~\ref{sec:arima} are
designed to do exactly that.  Note, though, that periodic patterns
appear only in signals that are at the low end of the complexity
spectrum.  If a signal contains little predictive structure, forecast
strategies like LMA and ARIMA have [[nothing to exploit and thus will
    be beaten out by]] [[needs wordsmithing from Liz]] the two simple
strategies described in this section.  This is explored further in
Sections~\ref{sec:accuracy} and~\ref{sec:results}.

@@

Perhaps the simplest way to capture and exploit the structure of data
is to fit a hyperplane to the dataset and then use that it to forecast
new data points.  The roots of this date back to Yule's 1927 invention
of the autoregressive schema~\cite{weigend93}, which forecasts the
next time step through a weighted average of past observations: $$px_i
= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
that is fit to the past $p+1$ [[???]] values of $x_i$ via the
backshift operator $B$ \footnote{I changed the superscript to $k$ to
  make this definition general---not connected to any of the terms
  that have specific meanings later in this section.}:
$$B^k x_i = x_{i-k}$$
%
with $k=1...p$ [[presumably?]].  To account for noise in the data, one
can add a so-called ``moving average'' term to the model:
$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
degree $q$ and $\epsilon$ is white noise with the same mean and
variance as the data.
% if you leave this in, you'll need to explain a lot about the data
% and the calculation
% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
To remove nonstationarities in the data, one can detrend it using a
differencing operation: $(1-B^d)x_i$.

A strategy that incorporates all three of these features is called a
\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
 $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%
{\color{red} You had superscripts on some of the $B$s and not on
  others.  I added them but I may have gotten them wrong.  I also
  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
  the first use of that construction.  Please make sure things are
  consistent, here and throughout this section.}  Here, $p$, $d$ and
$q$ correspond to the orders of the autoregressive, detrending, and
moving average terms, respectively.

If periodic structure is present in the data, a \emph{seasonal ARIMA}
model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%
$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
\Theta(B^Q)\theta(B^q)\epsilon_i$$
%
Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
and $Q$, respectively [[which accomplish what purpose?  why?  how?
    where did $D$ come from and what does it do?]].  $m$ is the
seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
    one should choose $P=Q=m$, since that's what the equation below
    implicitly does.]]  To use a seasonal ARIMA model, one first uses
the detrending term to remove nonstationarity: $$\hat{x_i} =
(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
\Theta(B^m)\theta(B^q)\epsilon_i$$

There is a vast amount of theory and literature regarding the
construction and use of this class of models; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear forecasting strategies.  In the sections that follow, we use
the automated fitting and forecasting techniques described
in~\cite{autoARIMA} to select model and differencing orders and fit
the corresponding polynomials to the time series.  This procedure uses
a combination of KPSS unit-root tests \cite{KPSSunit} and a
customization of the Canova-Hansen test \cite{Canova1995} to select
the differencing orders $d$ and $D$ and chooses the model order using
the Akaike information criterion (AIC) \cite{akaike1974}, conditioned
on the maximized likelihood of the model fitted to the detrended data,
$(1-B^m)^D(1-B^d) x_{i,obs}$.

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts will be inaccurate.
Another weakness of this method is prediction horizon, since an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To avoid this issue, we employ
it in a stepwise fashion: we build the model, perform a one-step
prediction, rebuild using the latest observations, and iterate until
the desired prediction horizon is reached.  (For consistency, we take
the same approach with the other three models in this study as well.)

>>>>

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $px_i$ at time $i$ is predicted using
the relation $$px_i = x_{i-1}$$ The \naive ~prediction strategy
averages the prior observations to generate the forecast: $$px_i =
\sum_{j=1}^{i-1}\frac{x_j}{i-1}$$ While both of these methods are
simplistic, they are not without merit.  For a time series near the
high end of the complexity spectrum---i.e., one that possesses very
little predictive structure---these two methods can actually be the
best choice.  In forecasting currency exchange rates, for instance,
sophisticated econometrics-based prediction models fail to
consistently outperform the random-walk method~\cite{rwMeese,rwCCE}.
These signals are volatile, noisy, and possess very little predictive
structure, but they do not---on the average---vary a great deal, so
the random-walk method's strategy of simply guessing the last known
value is not a bad choice.  If a signal has a unimodal distribution
with low variance, the \naive ~prediction strategy will perform quite
well---even if the signal is highly complex---simply because the mean
is a good approximation of the future behavior.  Moreover, a temporal
average effects a low-pass filtering operation.  In a signal with very
little predictive structure, then, the \naive ~prediction strategy
effectively mitigates the complexity.

Both of these methods have significant weaknesses, however.  Because
they do not model the temporal patterns in the data, or even the
distribution of its values, they cannot track changes in that
structure.  This causes them to fail in a number of important
situations.  Random-walk strategies are a particularly bad choice for
time series that change drastically at every time step.  In the worst
case---a large-amplitude square wave whose period is equivalent to
twice the sample time---a random-walk prediction would be exactly 180
degrees out of phase with the true continuation.  The \naive ~method
would be a better choice in this situation, since it would always
predict the mean.  It would, however, perform poorly when a signal has
a number of long-lived regimes that have significantly different
means.  In this situation, the inertia of the \naive ~method's
accumulating mean is a liability and the agility of the random-walk
method is an advantage, since it can respond quickly to regime shifts.

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The ARIMA and LMA
methods introduced in Sections~\ref{sec:lma} and~\ref{sec:arima} are
designed to do exactly that.  Note, though, that periodic patterns
appear only in signals that are at the low end of the complexity
spectrum.  If a signal contains little predictive structure, forecast
strategies like ARIMA and LMA have nothing to work with and thus will
often be outperformed by the two simple strategies described in this
section.  This is explored further in Sections~\ref{sec:accuracy}
and~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% here is the old version of this material, for use in other documents...
%
%Perhaps the simplest way to capture and exploit the structure of data
%is to fit a hyperplane to the dataset and then use that it to forecast
%new data points.  The roots of this date back to Yule's 1927 invention
%of the autoregressive schema~\cite{weigend93}, which forecasts the
%next time step through a weighted average of past observations: $$px_i
%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
%that is fit to the past $p+1$ [[???]] values of $x_i$ via the
%backshift operator $B$ \footnote{I changed the superscript to $k$ to
%  make this definition general---not connected to any of the terms
%  that have specific meanings later in this section.}:
%$$B^k x_i = x_{i-k}$$
%%
%with $k=1...p$.  To account for noise in the data, one can add a
%so-called ``moving average'' term to the model:
%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
%degree $q$ and $\epsilon$ is white noise with the same mean and
%variance as the data.
%% if you leave this in, you'll need to explain a lot about the data
%% and the calculation
%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
%To remove nonstationarities in the data, one can detrend it using a
%differencing operation: $(1-B^d)x_i$.  
%
%A strategy that incorporates all three of these features is called a
%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%%
%{\color{red} You had superscripts on some of the $B$s and not on
%  others.  I added them but I may have gotten them wrong.  I also
%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
%  the first use of that construction.  Please make sure things are
%  consistent, here and throughout this section.}  Here, $p$, $d$ and
%$q$ correspond to the orders of the autoregressive, detrending, and
%moving average terms, respectively.
%
%If periodic structure is present in the data, a \emph{seasonal ARIMA}
%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%%
%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
%\Theta(B^Q)\theta(B^q)\epsilon_i$$
%%
%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
%and $Q$, respectively [[which accomplish what purpose?  why?  how?
%    where did $D$ come from and what does it do?]].  $m$ is the
%seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
%    one should choose $P=Q=m$, since that's what the equation below
%    implicitly does.]]  To use a seasonal ARIMA model, one first uses
%the detrending term to remove nonstationarity: $$\hat{x_i} =
%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
%\Theta(B^m)\theta(B^q)\epsilon_i$$
% ...down to here
%%%%%%%%%%%%%%%%%%%%%%%%

Perhaps the simplest way to capture and exploit the structure of data
is to fit a hyperplane to the dataset and then use that it to make
predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$px_i = \sum a_i x_i$$ The weighting coefficients $a_i$
are computed using a polynomial of degree $p$ that is fit to the past
$p+1$ values of $x_i$.  To account for noise in the data, one can add
a so-called ``moving average'' term to the model; to remove
nonstationarities, one can detrend the data using a differencing
operation.  A strategy that incorporates all three of these features
is called a \emph{nonseasonal ARIMA model}.  If periodic structure is
present in the data, a \emph{seasonal ARIMA model}, which adds a
sampling operation that filters out periodicities, can be a good
choice.

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear forecasting strategies.  Fitting such a model to a dataset
involves choosing values for the various free parameters in the
autoregressive, detrending, moving average, and filtering
terms---e.g., the $p$ in the equation at the beginning of this
section.  We employ the automated fitting techniques described
in~\cite{autoARIMA} to accomplish this.  This procedure uses
sophisticated methods---KPSS unit-root tests~\cite{KPSSunit}, a
customization of the Canova-Hansen test~\cite{Canova1995}, and the
Akaike information criterion~\cite{akaike1974}, conditioned on the
maximized likelihood of the model fitted to the detrended data---to
select good values for the free parameters of the ARIMA model.

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts will be inaccurate.
Another weakness of this method is prediction horizon, since an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To avoid this issue, we build
forecasts in a stepwise fashion: i.e., fit the model to the existing
data, use it to perform a one-step prediction, rebuild the model using
the latest observations, and iterate until the desired prediction
horizon is reached.  (For consistency, we take the same approach with
the other three models in this study as well, even though that amounts
to artificially hobbling LMA.)
